---
title: "HW6 Linear mixed-effect models"
author: "Andrea Grillandi" 
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

## 0. Packages and data
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(lingtypology) # only for linguistic mapping
library(broom)
```

### UPSID database
In this dataset we have number of consonants and vowels in 402 languages collected from UPSID database (http://www.lapsyd.ddl.ish-lyon.cnrs.fr/lapsyd/).  
* `language` - language  
* `area` - language area according to Glottolog (http://glottolog.org/)  
* `consonants` - the number of consonants in the language  
* `vowels` - the number of vowels in the language  

```{r, warning=FALSE, message=FALSE}
upsid <- read_csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/upsid.csv")
upsid
```

```{r}
# you can map the languages using the lingtypology package
# map.feature(upsid$language,
#            features = upsid$area,
#            label = upsid$language,
#            label.hide = TRUE)
```

In this work, you will fit a number of linear and linear mixed-effect models that predict the number of vowels in a language by the number of consonants and other variables.

## 1. Linear model

### 1.1 
Make a scatterplot of the number of consonants and the number of vowels using the `ggplot()`.   
```{r 1.1, echo = FALSE}
p <- ggplot(upsid, aes(x=consonants, y=vowels))
p <- p + geom_point(color='darkblue') + theme_bw()
p
```

```{r}
# we use the theme_bw() theme, you can use other themes if you want
```
### 1.2
Fit the basic linear model `fit1` that predicts the number of vowels by the number of consonants. Look at the summary of the model.
```{r 1.2, include=FALSE}
fit1 <- lm(data = upsid, vowels ~ consonants)
summary(fit1)
```

### 1.3
Is `consonants` a significant predictor? Write down YES or NO.
```{1.3}
YES
# Because we see that p-value is less than 0.05, hence the two variables are intertwined.
```

### 1.4
*no evaluation*  
To draw predictions on the graph, we may use `fortify` function. It adds column entitled `.fitted` (note period) that is calculated as a prediction of a model as well as some other columns.

Try this code and see the result:

```{r}
head(fortify(fit1))
```
Another option is to use `augment` function from `broom` library that works in a similar way.

```{r}
head(augment(fit1))
```

### 1.5
Use one of these function to add a line to the scatterplot 1.1. You have to use `geom_line` and pass the result of `augment` or `fortify` as argument `data` in `geom_line`. (E.g. `geom_line(data = <result of augment>, ...`))

```{r 1.5, echo=FALSE}
p <- p + geom_line(data = fortify(fit1), aes(x = consonants, y = .fitted), color = "blue")
p
```

## 2. Mixed-effect models
Let us look at the data with respect to the `area` groups. 

### 2.1
Re-build the scatterplot `1.1` coloring the points by `area`.
```{r 2.1, echo=FALSE}
p2 <- ggplot(upsid, aes(x=consonants, y=vowels, color=area))
p2 <- p2 + geom_point() + theme_bw()# + stat_ellipse(type='norm', linetype=1)
p2
```

### 2.2
Use lmer() to fit the model `fit2` with random `area` group intercept. Your model is given by equation:

$$vowels = \beta_0 + \beta_1 \times consonants + u(area).$$
```{r 2.2, include=FALSE}
fit2 <- lmer(data=upsid, vowels ~ consonants + (1|area))
fortify(fit2)
```



### 2.3
Add the regression lines to the scatterplot `2.1` using `fortify` or `augment` methods.
```{r 2.3, echo=FALSE}
p2 <- p2 + geom_line(data = fortify(fit2), aes(x = consonants, y = .fitted, color = area))
p2
```

### 2.4
Interpret the results of the model. (Use `summary()` to get information about your model.) Is _consonansts_ variable still significant? How it changes after we added _area_ random effect? What can you say about lines on the graph? How are they located with respect to each other? Why?

```{r 2.4, include=FALSE}
summary(fit2)
```


```
Yes, it is still significant, yet p-value appears to be higher than before. The lines in the graph are all parallel one to the other, because the slope is determined by _consonants_ variable, with no correlation to the area, however they shift up and down according to the area (each group area presents different intercepts).
```

## 3. Mixed-effect models with random slopes.
### 3.1
Fit the model `fit3` with random slope that depends on _area_. Your model is given by formula:

$$vowels = \beta_0 + (\beta_1 + u(area))\times consonants.$$
You can consult [this](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#model-specification) manual to choose the correct syntax in your `lmer` formula.

```{r 3.1, include=FALSE}
fit3 <- lmer(data=upsid, vowels ~ consonants + (1+consonants|area))
```

### 3.2
Draw a figure with prediction lines:

```{r 3.2, echo=FALSE}
p3 <- p2 + geom_line(data = fortify(fit3), aes(x = consonants, y = .fitted, color = area))
p3
```

### 3.3
Interpret the results. Is _constants_ still significant? Why? What can you say about lines on the graph? How are they located with respect to each other? Why?

```{r 3.3, include=FALSE}
summary(fit3)
```

```
In the third model, the variable consonants lost significance (look at p-value). We can note that different lines have different slopes, because area variable influences slope calculation (we are assuming a mixed-effect between area and number of consonants).
```

## 4. Mixed-effect models with random intercept and random slopes.
Now let us assume we have both random intercept and random slope (and they are not correlated). Our model is of the form:

$$vowels = \beta_0 + u_1(area) + (\beta_1 + u_2(area))\times consonants.$$

### 4.1
Fit the model `fit4`.

```{r 4.1, include=FALSE}
fit4 <- lmer(data=upsid, vowels ~ consonants + (1|area) + (1+consonants|area))
```

### 4.2
Draw a figure with predictions:  

```{r 4.2, echo=FALSE}
p4 <- p2 + geom_line(data = fortify(fit4), aes(x = consonants, y = .fitted, color = area))
p4
```

### 4.3
Interpret the results. Is _constants_ significant? How can we interpret the difference between the prediction lines of these three models?

```{r 4.3}
summary(fit4)
```

```
Also regarding fit4, consonants variable has small significance. The difference among the 3 models is due to the different components we take into consideration when calculating the regression. If we assume no mixed-effect between area and number of consonants, we see how different areas present dissimilarities only in intercepts (fit2), meaning we are able to observe only a difference in distribution; however, in fit2 and fit3, we can observe the interaction bewtween the variables number of consonants and area, since slope is determined by their mixed-effect (hence, we are assuming there is a mixed-effect).
```
